\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-14 docs/parco13/abstract.tex
Some minor grammatical changes;  added bibliography info (see ../Biblio/amr.bib), several comments included in the text.

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

@@

% Two or more authors:
%\author[A]{\fnms{} \snm{}},
%\author[B]{\fnms{} \snm{}}
%\runningauthor{}
%\address[A]{}
%\address[B]{}
%
\author[A]{\fnms{Carsten} \snm{Burstedde}%
\thanks{Corresponding author.  E-mail: \texttt{burstedde@ins.uni-bonn.de}}},
\author[B]{\fnms{Donna} \snm{Calhoun}},
\author[C]{\fnms{Kyle} \snm{Mandli}} and
\author[C]{\fnms{Andy R.} \snm{Terrel}}
\runningauthor{C.\ Burstedde et.al.}
\address[A]{Institut f\"ur Numerische Simulation, Universit\"at Bonn, Germany}
\address[B]{Boise State University, Idaho, USA}
\address[C]{Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin, USA}

\begin{abstract}
%
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ of dimensions $m^d$, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to work with the \pforest AMR library since it allows to compose the mesh
from multiple mapped octrees, thus enabling the cubed sphere and other
nontrivial multiblock geometries.
%
\end{abstract}

% \begin{keyword}
% adaptive mesh refinement,
% clawpack,
% HPC,
% manycore
% \end{keyword}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers that use
adaptive mesh refinement, especially when working with unstructured meshes.  Of
course, it has been realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as it has been done with high-order
spectral element \cite{TufoFischer99}, finite volume
\todo{add favorite citation}, or discontinuous Galerkin
\cite{BursteddeGhattasGurnisEtAl10} methods.
%
% For example, SEs optimized for computational speed have originally been
% implemented on unstructured conforming meshes \cite{TufoFischer99} and later
% been extended to non-conforming adaptive meshes \cite{FischerKruseLoth02,
% BursteddeGhattasGurnisEtAl10}.

To enable hardware acceleration for parallel dynamic AMR, we like to build upon
the forest-of-octrees paradigm because of its low overhead and proven
scalability \cite{BursteddeWilcoxGhattas11}.  This approach identifies each
octree leaf with a mesh element.  In this work, we go
beyond the traditional high-order element and define each element to be a
dense compute patch with $m^d$ DOFs.  In fact, this approach resembles
block-structured AMR \cite{ColellaGravesKeenEtAl07}
\todo{favorite BoxLib citation}
except that the patches are not overlapping, which enables us to capitalize on
our previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The Clawpack software \cite{LeVeque97}
provides a popular implementation of such a patch.  It has
been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR \todo{AMRClaw
citation}.

In this paper we describe how we design the coupling of forest-of-octree AMR
with Clawpack at the leaf level.  We comment on challenges that arise in
enabling multiblock geometries and efficient parallelism and close with a range
of numerical examples that demonstrate the conceptual improvements in relation
to other approaches.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
hithertwo disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (24 bytes per leaf) is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).
\end{enumerate}

% yields periodicity as a special case
A particular feature of \forestclaw is that the generic handling of multiblock
geometries is inherited from \pforest, identifying each octree as a block, and
each leaf as a patch.  Each block is understood as a reference unit cube with
its own geometric mapping.  The connectivity of the blocks is static and can be
created by external hexahedral mesh generators, elimintating the need to encode
it by hand.

A main challenge is presented by the fact that the patch neighborhood is only
known to \pforest, from which it needs to be propagated to the
numerical code in \forestclaw that implements the interaction with ghost
patches.  To this end, we define an interface that allows distributed read-only
access to the sequence of blocks, the list of patches for each, and a lookup of
neighbor patches (and their relative orientation which is nontrivial between
blocks).  The exchange of ghost data and parallel partitioning is provided
transparently by \pforest.  Mesh modification directives, such as local
refinement and coarsening, are called from \forestclaw and relayed to \pforest.

@@

\forestclaw uses iterators over all leafs, optionally restricted to a given
level.  Random access is possible and used when executing $O(1)$ time neighbor
lookups.  Looping over the patches in the order prescribed by the forest and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse.
% since the ghost patches are likely closeby with respect to the $z$-order.

@@

%%%%%%%%%%% The bibliography starts:
\bibliographystyle{unsrt}
\bibliography{ccgo}

>>>>

\newcommand{\comment}[1]{\textcolor{green}{[DAC: #1]}\xspace}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

@@

% Two or more authors:
%\author[A]{\fnms{} \snm{}},
%\author[B]{\fnms{} \snm{}}
%\runningauthor{}
%\address[A]{}
%\address[B]{}
%
\author[A]{\fnms{Carsten} \snm{Burstedde}%
\thanks{Corresponding author.  E-mail: \texttt{burstedde@ins.uni-bonn.de}}},
\author[B]{\fnms{Donna} \snm{Calhoun}},
\author[C]{\fnms{Kyle} \snm{Mandli}} and
\author[C]{\fnms{Andy R.} \snm{Terrel}}
\runningauthor{C.\ Burstedde et al.}
\address[A]{Institut f\"ur Numerische Simulation, Universit\"at Bonn, Germany}
\address[B]{Boise State University, Idaho, USA}
\address[C]{Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin, USA}

\begin{abstract}
%
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ of dimensions $m^d$, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to work with the \pforest AMR library since it allows us to compose the mesh
from multiple mapped octrees and enable the cubed sphere and other
nontrivial multiblock geometries.
%
\end{abstract}

% \begin{keyword}
% adaptive mesh refinement,
% wave propagation algorithms
% clawpack,
% HPC,
% manycore
% \end{keyword}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement, especially when working with
unstructured meshes \comment{This sounds we are now going to talk
  about unstructured meshes.}.  Of course, it has been realized early
that it helps to aggregate degrees of freedom (DOF) at the element
level, as it has been done with high-order spectral element
\cite{TufoFischer99}, finite volume
\cite{paramesh, chombo, samrai}, or discontinuous Galerkin
\cite{BursteddeGhattasGurnisEtAl10} methods.
\comment{Are you looking for references to parallel AMR codes?  Or more general AMR codes based on the FVM? }

% For example, SEs optimized for computational speed have originally been
% implemented on unstructured conforming meshes \cite{TufoFischer99} and later
% been extended to non-conforming adaptive meshes \cite{FischerKruseLoth02,
% BursteddeGhattasGurnisEtAl10}.

To enable hardware acceleration for parallel dynamic AMR, we like to
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In this work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989, be-le:1991,
ColellaGravesKeenEtAl07} except that the
patches are not overlapping, which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The Clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

In this paper we describe our design for the coupling of forest-of-octree AMR
with Clawpack at the leaf level.  We comment on challenges that arise in
enabling multiblock geometries and efficient parallelism and conclude with a range
of numerical examples that demonstrate the conceptual improvements in relation
to other approaches.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
\comment{I am not sure about the Python dependency;  the latest version of Clawpack relies on
Python to generate an input file, but the code itself is F77 (or maybe some F90). }
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (approx 100 bytes per leaf
\comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}) is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).
\end{enumerate}

% yields periodicity as a special case
A particular feature of \forestclaw is that the generic handling of multiblock
geometries is inherited from \pforest, identifying each octree as a block, and
each leaf as a patch.  Each block is understood as a reference unit cube with
its own geometric mapping.  The connectivity of the blocks is static and can be
created by external hexahedral mesh generators, eliminating the need to encode
it by hand.

A main challenge is presented by the fact that the patch neighborhood
is only known to \pforest.  This patch connectivity information needs
to be propagated to the numerical code in \forestclaw that implements
the interaction with neighbor patches via the use of a layer of ghost
cells surrounding each patch.  To this end, we define an interface
that allows distributed read-only access to the sequence of blocks,
the list of patches for each, and a lookup of neighbor patches (and
their relative orientation which is nontrivial between blocks).  The
exchange of ghost data and parallel partitioning is provided
transparently by \pforest.  Mesh modification directives, such as
local refinement and coarsening, are called from \forestclaw and
relayed to \pforest.

@@

\forestclaw uses iterators over all leafs, optionally restricted to a given
level.  Random access is possible and used when executing $O(1)$ time neighbor
lookups.  Looping over the patches in the order prescribed by the forest \comment{tree?} and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse.
% since the ghost patches are likely closeby with respect to the $z$-order.

@@

%%%%%%%%%%% The bibliography starts:
\bibliographystyle{unsrt}
\bibliography{ccgo,../Biblio/amr}
